### 2021-10-25 documents/sections/extensions.tex
fixed proto-pekodian 'to go'

For \arara, \textcite{alves2017arara} reports six\footnote{Seven under her analysis, which treats the two meanings of \obj{it͡ʃi} \qu{to be, to lie down} as different verbs.} \gl{s_a_} verbs which have a first person marker \obj{w(ɨ)-} rather than \obj{k-}, listed in \exref{arairr}.
We take these \arara verbs as a starting point for the other Pekodian languages, as \textcite{alves2017arara} gives the most detailed description of person markers for any language of the branch.

@@

\input{floats/aku1sa.tex}

>>>>

For \arara, \textcite{alves2017arara} reports six\footnote{Seven under her analysis, which treats the two meanings of \obj{it͡ʃi} \qu{to be, to lie down} as different verbs.} \gl{s_a_} verbs which have a first person marker \obj{w(ɨ)-} rather than \obj{k-}, listed in \exref{arairr}.
%We take these \arara verbs as a starting point for the other Pekodian languages, as \textcite{alves2017arara} gives the most detailed description of person markers for any language of the branch.

@@

\input{floats/aku1sa}

%Let us first consider a scenario where \obj{k-} was the older \gl{1}\gl{s_a_} marker, occurring on both \obj{e}- and \obj{ə}-initial verbs.
%Next, \obj{t-}/\obj{t͡ʃ-} would have come in and replaced \obj{k-} before \obj{e}, but (mostly) not before \obj{ə}, resulting in the observable distribution.
%This scenario is not very satisfactory, since \obj{k-} is replaced in one completely unmotivated phonological environment, but not in the other.
%
%The other scenario is that \rc{t-} was the old first person marker occurring on the vast majority of \gl{s_a_} verbs.
%Subsequently, it developed into \rc{t͡ʃ-} / \_\obj{e}, but remained \rc{t-} / \_\obj{ə}; \obj{t͡ʃ}\goodtilde\obj{ʃ}\goodtilde\obj{s} and \obj{t} are phonemically contrastive in \akuriyo \parencite[16]{meira1998proto}.
%Thus, \obj{k-} can be seen as replacing \obj{t-}, but not \obj{t͡ʃ-}.
%The occurrences of \obj{t-} / \_\obj{ə} mentioned by \citeauthor{meira1998proto} are then remnants of the earlier prevalence of \obj{t-}.
%While this scenario also lacks a clear motivation, it at least explains the largely phonologically complementary distribution of \obj{k-} and \obj{t-} by the fact that the marker which was replaced had a phonemically different form than the one which was preserved.
%The other scenario which sees \rc{k-} as older would require \rc{t-} replacing \obj{k-} only in a specific environment.\footnote{Note that this process would be plausible even if the instances of \obj{t-} / \_\obj{ə} were due to \trio influence, as suggested by \textcite[108]{meira1998proto} -- these would then be a later reintroduction, replacing \obj{k-}, instead of remnants of the introduction of it.}

It has been speculated that the \trio \gl{s_a_} verb with a first person \obj{k-} marker, \obj{weka}/\obj{oeka} \qu{to defecate} might have something to do with \akuriyo \obj{k-}, which would potentially make the innovation of \rc{k-} a \PTir matter \parencite[116]{meira1998proto}.
However, this hypothesis faces a problem in that the first person form for \akuriyo \qu{to defecate} is \obj{j-ereina-} \parencite[88]{gildea1994akuriyo}.
Another hypothesis is that it originated in the corresponding \kaxui form \obj{ku-weka-} (form provided by Spike Gildea, p.c.), although the occurrence of \obj{o} in \trio would still need explanation.
%Rather, it seems likely that this irregular first person form originated in a class switch from \gl{s_p_} to \gl{s_a_} (\Cref{sec:resistant_verbs}), although it is unclear how exactly that switch resulted in the irregular form.

\subsection{\carijo \obj{j-}}
\label{sec:carijo}
As discussed by \textcite[105--107]{meira1998proto}, \carijo has extended the \gl{1}\gl{s_p_} marker \obj{j(i)-} to \gl{s_a_} verbs.\footnote{Since almost all \gl{s_a_} verbs are V-initial, only the \envr{}{V} allomorph \obj{j-} is attested.}
This, in combination with the extension of \gl{2}\gl{s_a_} \obj{m-} and \gl{1+2}\gl{s_a_} \obj{kɨt-}/\obj{kɨs-} to \gl{s_p_} verbs, results in a single unified \gl{s} category \exref{carpar}.

\ex<carpar> \carijo \parencites[106]{meira1998proto}[173]{robayo2000avance}\\
\begin{tabular}[t]{@{}lll@{}}
& \obj{tuda} \qu{to arrive} & \obj{eharaga} \qu{to dance}\\
\gl{1} & \obj{ji-tuda-} & \obj{j-eharaga-}\\
\gl{2} & \obj{mɨ-tuda-} & \obj{m-eharaga-}\\
\gl{1+2} & \obj{kɨsi-tuda-} & \obj{kɨs-eharaga-}\\
\gl{3} & \obj{ni-tuda-} & \obj{n-eharaga-}\\
\end{tabular}
\xe
%
Although the split-\gl{s} system has been lost entirely, former \gl{s_a_} verbs can be identified by taking a detransitivizer, like \obj{ehɨnəhɨ} \qu{to fight}, derived from \obj{hɨnəhɨ} \qu{to kill} \parencite[179]{robayo2000avance} \exref{car-30}.
%The same marker is also found in \gl{3}>\gl{1} scenarios \exref{car1p.car-1} and possessed nouns \exref{car1p.car-31}.

\ex<car-30>\carijo \parencite[][79]{koch1908hiana}\\
\begingl
\glpreamble hẹnẹ(x)tónoko-mā́ɽẹ y-e̱-hẹnẹ́(x)yai̯//
\gla hɨnəhtono-ko=marə j-e-hɨnəhɨ-jai//
\glb enemy-\gl{pl}=with \gl{1}-\gl{detrz}-kill-\gl{npst}.\gl{cert}//
\glft \qu{I fight with the enemies.}//
\endgl
\xe
%

%\pex<car1p>\carijo \parencite[][59, 53]{guerrero2016carijo}
%\begin{multicols}{2}
%\a<car-1>
%\begingl
%\gla jɨ-hɨnəh-ɨ//
%\glb \gl{3}>\gl{1}-kill-\gl{pfv}//
%\glft \qu{S/he killed me.}//
%\endgl
%\a<car-31>
%\begingl
%\gla j-owo-rɨ//
%\glb \gl{1}\gl{poss}-uncle-\gl{pert}//
%\glft \qu{my uncle}//
%\endgl
%\end{multicols}
%\xe
%

As noted in \cref{sec:taranoan}, this extension also erased any traces of a potential \PTar \gl{1}\gl{s_a_} marker \rc{t-}.
However, it did not fully eclipse the old \gl{1}\gl{s_a_} marker \rc{w-}, which is attested as being preserved in the verbs \obj{tə} \qu{to go} \exref{car1w.car-24} and \obj{a} \qu{to be} \exref{car1w.car-25}.

\pex<car1w>\carijo \parencite[][5, 42]{guerrero2016karihona}
\a<car-24>
\begingl
\gla wɨ-tə-e=rehe//
\glb \gl{1}-go-\gl{npst}=\gl{frust}//
\glft \qu{I almost go (but I am not going to go).}//
\endgl
\a<car-25>
\begingl
\gla əji-marə-ne w-a-e//
\glb \gl{2}-with-\gl{pl} \gl{1}-be-\gl{npst}//
\glft \qu{I am with you all.}//
\endgl
\xe
%
%Interestingly enough, the verb \qu{to be} only shows \obj{w-} in its \obj{a} root allomorph, but not in the \obj{et͡ʃi} root allomorph, where it has \obj{j-} \exref{car1y.mivida-12}.
%In a somewhat foreshadowing move, I will note that \PTar \rc{(ət)epɨ} \qu{to come} was also affected by the \carijo extension, as seen in \exref{car1y.car-18}.
%As discussed in \cref{sec:taranoan}, this verb preserves \rc{w-} in \akuriyo and \trio \exref{tri-105}.
%
%\pex<car1y>\carijo
%\a<mivida-12>
%\begingl
%\glpreamble iretibə et͡ʃinəme gərə jet͡ʃiɨ//
%\gla ireti-bə et͡ʃi-nə=me gərə j-et͡ʃi-ɨ//
%\glb then-from be-\gl{inf}=\gl{attrz} still \gl{1}-be-\gl{pfv}//
%\glft \qu{Then I was already grown up.} \parencite[][177]{robayo1989rame}//
%\endgl
%\a<car-18>
%\begingl
%\gla əji-wa-e j-eh-ɨ//
%\glb \gl{2}-search-\gl{sup} \gl{1}-come-\gl{pfv}//
%\glft \qu{I came looking for you.} \parencite[][102]{guerrero2019carijo}//
%\endgl
%\xe
%
%\ex<tri-105>\trio \parencite[][337]{triocarlin2004}\\
%\begingl
%\gla ene w-əe//
%\glb see.\gl{sup} \gl{1}\gl{s_a_}-come//
%\glft \qu{I came to see him/her.}//
%\endgl
%\xe

\subsection{\yukpa \obj{j-}}
\label{sec:yukpa}
\yukpa has lost most of its \setone constructions, mostly employing innovative ones \parencite{meira2006syntactic}.
It preserves the \setone prefixes in the immediate past, where the \gl{1}\gl{s_p_} marker appears to have been extended to \gl{s_a_} verbs, as evidenced by the paradigms in \exref{yukpaintr}.
The loss of split-\gl{s} in favor of the \gl{1}\gl{s_p_} marker is accompanied by the \gl{2}\gl{s_a_} marker \obj{m(ɨ)-} displacing \gl{2}\gl{s_p_} \rc{ə(j)-}, and \gl{1+2} being lost as an inflectional value.
%It has lost the split-\gl{s} system in general

\ex<yukpaintr> \yukpa \parencites[72, 76]{largo2011yukpa}[139]{meira2006syntactic}\\
\begin{tabular}[t]{@{}llll@{}}
& \obj{otum} \qu{to wash self} & \obj{nɨ} \qu{to sleep} & \obj{ata} \qu{to fall}\\
\gl{1} & \obj{j-otum-} & \obj{jɨ-nɨ-} & \obj{j-ata-}\\
\gl{2} & \obj{m-otum-} & \obj{mɨ-nɨ-} & \obj{m-ata-}\\
\gl{3} & \obj{n-otum-} & \obj{nɨ-nɨ-} & \obj{n-ata-}\\
\end{tabular}
\xe
%
The normal intransitive first person marker is \obj{j(ɨ)-}, including those with clear reflexes of \detrz, like \obj{otum} \qu{to wash self} in \exref{yukpaintr}.
The analysis that this is a reflex of the \gl{1}\gl{s_p_} marker \rc{u(j)-} is supported by the same form occurring in \gl{3}>\gl{1} scenarios \exref{yukpatr.yuk-5}.

\pex<yukpatr>\yukpa \parencite[][139]{meira2006syntactic}
\a<yuk-5>
\begingl
\gla aw j-esare//
\glb \gl{1}\gl{pro} \gl{3}>\gl{1}-see//
\glft \qu{S/he saw me.}//
\endgl
\a<yuk-1>
\begingl
\gla aw {\normalfont ∅}-esare//
\glb \gl{1}\gl{pro} \gl{1}>\gl{3}-see//
\glft \qu{I saw it.}//
\endgl
\xe
On the other hand, \gl{1}>\gl{3} scenarios are zero-marked \exref{yukpatr.yuk-1}.
Since \PC \gl{1}\gl{s_a_} \rc{w(ɨ)-} was extended to \gl{1}>\gl{3} scenarios in most languages \parencite[81--82]{gildea1998}, and since it is prone to phonological erosion across the family \pcref{sec:pekodian,sec:waiwaian,sec:akuriyo}, we argue that the zero in \gl{1}>\gl{3} scenarios is in fact the \yukpa reflex of \rc{w-}.
The same zero marking is attested for a single intransitive verb: \obj{to} \qu{to go} \exref{yuk-7}.
Based on regular C-initial verbs taking \obj{jɨ-}, like \qu{to sleep} in \exref{yukpaintr}, it is clear that also C-initial \obj{to} \qu{to go} preserved the old \gl{1}\gl{s_a_} marker \rc{w(ɨ)-}, which was lost due to phonological erosion.

\ex<yuk-7>\yukpa \parencite[][139]{meira2006syntactic}\\
\begingl
\gla aw {\normalfont ∅}-to//
\glb \gl{1}\gl{pro} \gl{1}\gl{s_a_}-go//
\glft \qu{I went.}//
\endgl
\xe
